---
title: "Mental Health and its Impact on Bullying in Schools"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    #df_print: paged
---
# Abstract

  While mental health has always been a factor in our education, it has only been recently that it has risen to the forefront of conversation about the impact it has on the education and school environments. In order to capture the impact mental health has on our educational systems today, we look at the mental health of several children in US high school level ages.

  We seek to understand what most commonly plagues teenagers in the mental health spectrum and what we can do to help address and alleviate this issue.

# Initial Setup
```{r}
rm(list = ls())
```

## Libraries
```{r}
library(sas7bdat)
library(tidyverse)
library(ggplot2)
library(data.table)
library(mosaic)
library(ggformula)
library(MASS)
library(dplyr)
library(sjstats)
library(pwr)
library(grid)
library(gtable)
```

## Datasets
```{r}
schooldataraw <- read.sas7bdat("C:\\Users\\julia\\OneDrive\\Desktop\\2018109\\07_FINAL PU DATA_February 5, 2018\\pu_ssocs16.sas7bdat")
studenthealthraw <- read.csv("C:\\Users\\julia\\Downloads\\DASH_-_Global_School-based_Student_Health_Survey__GSHS_.csv")
```

## Functions
```{r}
convertBoolean <- function(x) {
  return(x==1)
}
```

## Palettes
```{r}
threeColorPalette <- c("red3", "gold", "forestgreen")
fiveColorPalette <- c("red3", "orange", "gold", "forestgreen", "royalblue")
sevenColorPalette <- c("red3", "orange", "gold", "forestgreen", "royalblue", "purple2", "pink1")
```

# Data Wrangling and Cleaning

## Cleaning the 2018-2019 School Survey on Crime and Safety
```{r}
#selecting and cleaning up the chosen variables
schooldata <-
  schooldataraw %>%
  dplyr::select(C0174, C0662, C0664, C0666, C0668, C0670, C0672, C0674, C0676, C0678, C0670, C0672, C0674, C0676, C0678, C0680, C0682, C0684, C0686, C0265, C0267, C0273, C0376) %>%
  transmute(id = row_number(), antiBullying = C0174,
         assessAva = 6 - (C0662 + C0664 + C0666),
         treatAva = 6 - (C0668 + C0670 + C0672),
         limitMHProfAccess = C0674, limitFunding = C0676, limitLegal = C0678, limitParental = C0680, 
         limitComm = C0682, limitPaymentPol = C0684, limitStigma = C0686,
         discCyber = C0265, discBullying = C0267, recogBullying = C0273,
         bullyingRate = C0376)
```

```{r}
booleanRows = c(2, 12, 13, 14)

boolSchoolData <-
  lapply(schooldata[,booleanRows], convertBoolean)

schooldata[ , 2  ] <- boolSchoolData[1]
schooldata[ , 12 ] <- boolSchoolData[2]
schooldata[ , 13 ] <- boolSchoolData[3]
schooldata[ , 14 ] <- boolSchoolData[4]
```

```{r}
limitations <- data.table(comm = schooldata$limitComm, 
                          funding = schooldata$limitFunding, 
                          legal = schooldata$limitLegal, 
                          MHProfAccess = schooldata$limitMHProfAccess, 
                          parental = schooldata$limitParental, 
                          paymentPol = schooldata$limitPaymentPol, 
                          stigma = schooldata$limitStigma)
```

## Cleaning the Global Student Health Survey
```{r}
studenthealth <-
  studenthealthraw %>%
  transmute(Topic, Description, Question = Greater_Risk_Question, RespPct = Greater_Risk_Data_Value, Sample_Size, Sex, Age) %>%
  filter(Topic == "Mental Health" & Sex != "Total" & 
           Age != "Total" & Age != "13-15" & Age != "13-17") %>%
  na.omit()

studenthealth <-
  studenthealth %>%
  group_by(Age, Sex, Question) %>%
  mutate(totalSS = sum(Sample_Size)) %>%
  mutate(AffRespPct = sum((Sample_Size)*(RespPct))/totalSS) %>%
  dplyr::select(Sex, Age, Question, AffRespPct, totalSS) %>% 
  unique() %>%
  arrange(Sex, Age)
```

# Codebooks

## 2018-2019 School Survey on Crime and Safety
| Variable Name | Description |
| --- | --- |
| ID | The ID number of the School, given for purposes of the study |
| antiBullying | A boolean describing whether a school has anti-bullying programs put into place. |
| assessAva | An aggregate describing whether a school has mental health services for diagnostic assessment available, with a score of 3 being the highest. |
| treatAva | An aggregate describing whether a school has mental health services for treatment available, with a score of 3 being the highest. |
| limitMHProfAccess | A score describing whether inadequate access to mental health professionals was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitFunding | A score describing whether inadequate funding was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitLegal | A score describing whether potential legal issues, such as malpractice, or insufficient supervision were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitParental | A score describing whether lack of parental support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitComm | A score describing whether lack of community support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitPaymentPol | A score describing whether payment polcies were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitStigma | A score describing whether concern for potential stigmatization of the student was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| discCyber | A boolean describing whether school staff were trained in discipline policies and practices related to cyberbullying. |
| discBullying | A boolean describing whether school staff were trained in discipline policies and practices related to bullying. |
| recogBullying | A boolean describing whether school staff were trained in recognizing bullying. |
| bullyingRate | A score describing how often bullying occurs at a school, with 1 being daily and 5 being never. |

## Global Student Health Survey
| Variable Name | Description |
| --- | --- |
| Topic | A variable meant for categorizing, used to filter down to just mental health. |
| Question | A question asked related to mental health, with the relevant time frame being the 12 months before the survey |
| AffRespPct | Percent of people who responded yes to the question. |
| Sample Size | The amount of people who answered this question, with the specified sex and age. |
| Sex | The sex of the recipients. |
| Age | The age of the recipients. |

# Data
```{r}
schooldata
studenthealth
```


# How Effective Are Antibullying Programs
  One of the most commonplace programs put into schools to combat bullying, are the aptly named anti-bullying programs, where teachers are trained to recognize bullying, as well as policies and practices regarding both cyber and physical bullying. We also look at the general response of "anti-bullying" where a school responds affirmatively if there are any anti-bullying programs put into place. 

```{r fig.height=7, fig.width=8}
schoolgraphdata1 <-  
  schooldata %>%
  group_by(discBullying, discCyber) %>%
  summarize(meanRate = mean(bullyingRate))

dummyVerts <- data.frame(discCyber = c(T, F, T, F), discBullying = c(T, T, F, F), Z = schoolgraphdata1$meanRate)

sgraph1 <-
  schooldata %>%
  ggplot(aes(x = bullyingRate)) +
  geom_density(aes(x = bullyingRate, fill = recogBullying, alpha = 0.05)) +
  facet_grid(rows = vars(discBullying), cols = vars(discCyber)) +
  geom_vline(data = dummyVerts, aes(xintercept = Z))

sgraph1 + 
  labs(x = "Trained on Combating Cyberbullying", 
       y = "Trained on Combating Bullying", 
       title = "Bullying Rates With Antibullying Programs Put Into Place") + 
  theme(axis.title.x = element_text(vjust = 205),
        axis.title.y = element_text(angle = 270, vjust = 207))
```

## Bullying Rates for Various Combinations of Anti-Bullying Programs

```{r}
sgraph2 <-
  schooldata %>%
  ggplot(aes(x = bullyingRate, y = recogBullying, color = factor(antiBullying)))

sgraph2 +
  geom_boxplot(lwd = 0.3, color = "black") +
  geom_jitter(alpha = .15) +
  facet_grid(discBullying ~ discCyber) +
  theme(axis.text.x = element_text(vjust = 1))
```


```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ recogBullying + antiBullying + discCyber + discBullying, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 observations
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```

## Breakdown of the Effectiveness of Anti-Bullying Programs
  After running an ordered logistic regression model, we can see that none of these p values for our predictors are particularly significant, suggesting that training school staff in anti-bullying programs, recognizing bullying, and training them to learn discipline policies for both bullying and cyberbullying are all ultimately ineffective, and these policies don't create a significant impact on bullying rates.

  Therefore, if all of these commonplace school anti-bullying programs are ineffective, what programs can schools put into place to create a significant drop in bullying rates? A good place to start looking is seeing what schools feel limit them from helping students out on the mental health spectrum.

  Sidenote: All of the differences between our levels for bullyingRates are significant, especially for the extremes, 1 and 5, which is reassuring.

## Which limitations are most common to schools?
```{r}
limitationsLong <- pivot_longer(limitations, everything(), names_to = "limType")

limitationsLong %>%
  group_by(limType, value) %>%
  summarize(valCount = n()) %>%
  ggplot(aes(x = limType, y = valCount)) + 
  geom_col(aes(x = limType, y = valCount, fill = as.character(value))) +
  scale_fill_manual(values = threeColorPalette)
```

```{r}
limBully = data.table(limitations, schooldata$bullyingRate)
limBully <-
  limBully %>%
  rename(bullyingRate = V2)

limitationsLong2 <- pivot_longer(limBully, cols = c("comm", "funding", "legal", "MHProfAccess", "parental", "paymentPol", "stigma"), names_to = "limType") %>%
  transmute(limType = limType, limScore = value, bullyingRate = bullyingRate) 

limitationsLong3 <-
  limitationsLong2 %>%
  group_by(limType, limScore, bullyingRate) %>%
  summarize(bullyingRateCount = n())
```

```{r}
limitationsLong2 %>%  
  ggplot() +
  aes(x = limScore, fill = limType) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sevenColorPalette)
```

## Breakdown of What Do Schools Feel They Need and What Do Schools Actually Need

  Although, we might say some combinations like Funding being a very common highly limiting factor jump out to us immediately, we should confirm that it does in fact have a significant impact on bullying rates at schools, as well as any others which might have a noticeable impact. We do this through taking an analysis of variance of the different mean bullying rates for different limitation types in order to see which limitation type has the most impact on bullying rates.
  
  We do this by looking at different combinations of limitation score and limitation type. We plan to do this to answer the following key research question.
 
### How effective are different limitations/resources at combating bullying in schools?
```{r}
limTypes = c("comm", "funding", "legal", "MHProfAccess", "parental", "paymentPol", "stigma")
limPVals = c()
limCohens = c()

for(type in limTypes) {
  limitationsLongTemp <-
    limitationsLong2 %>%
    filter(limType == type)
  
   pVal <- summary(aov(bullyingRate ~ limScore, data = limitationsLongTemp))[[1]][["Pr(>F)"]][[1]]
   limPVals = c(limPVals, pVal)
   
   limCohen <- cohens_f(aov(bullyingRate ~ limScore, data = limitationsLongTemp))[1,2]
   limCohens = c(limCohens, limCohen)
}

data.table(limTypes, limPVals, limCohens)
```

  We see that each limitation type does have an effect on a school's bullying rate, with the level of limitation effecting a school's bullying rate. However, we must look at whether the effects of these limitations are large enough to matter. 
  
  Using Cohen's F statistic, we get Cohen's F values in a range from about 0.077 to 0.127. According to Cohen's suggested Cohen's F Values of 0.1, 0.25, and 0.4 of small, medium, and large effect sizes respectively, we see that all of our effect sizes are in the "small" range, suggesting that these limitations have not much effect on bullying rates, despite their significance.
  
  We can see a graph of how bullying rates differ at different combinations of limitation type and level below.

## Straight Counts of Occurences of Bullying Rates at Various Levels of Specific Limitations
```{r}
limitationsLong3 %>%  
  ggplot() +
  aes(x = limType, fill = as.character(bullyingRate), y = bullyingRateCount) +
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) +
  facet_wrap( ~ limScore) +
  scale_fill_manual(values = fiveColorPalette)
```

## Straight Counts of Occurences of Bullying Rates at Various Levels of Specific Limitations
```{r}
temp <-
  limitationsLong3 %>%
  group_by(limType, limScore) %>%
  summarize(totCount = sum(bullyingRateCount))

limitationsLong3 %>%
  left_join(temp, by = c("limType", "limScore")) %>%
  mutate(propRate = bullyingRateCount/totCount) %>%
  ggplot() +
  aes(x = limType, fill = as.character(bullyingRate), y = propRate) +
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) +
  facet_wrap( ~ limScore) +
  scale_fill_manual(values = fiveColorPalette)
```

  As we can see, although we see what could be a very slight increase in the proportion of schools with lower bullying rate at higher limitation levels, we see it's not a very large and distinct increase, mirroring what Cohen's F statistic suggests. 

  Clearly, several of the most common anti-bullying programs and styles are all ineffectual, and the common limitations that schools frequently cite turn out not to have too much impact at all. This seems to spell out a grim future for the next generation of students, with bullying remaining a prevalent problem throughout all of our schools, with all of the most common methods against them being ultimately ineffectual and our schools asking for resources with virtually no impact. 
  
  Therefore, where should we go to find a solution to the bullying epidemic in our schools? Following our link between mental health and bullying, we move on to:
  
# Mental Health & Bullying: How Does Assessment and Treatment of Mental Health Impact Bullying
```{r}
sgraph3 <-
  schooldata %>%
  ggplot(aes(x = bullyingRate)) +
  geom_density(aes(x = bullyingRate)) +
  geom_rect(fill = "blue", alpha = 0.003, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf)) +
  facet_grid(rows = vars(assessAva), cols = vars(treatAva), labeller = label_both)

temp <-
  schooldata %>%
  group_by(assessAva, treatAva) %>%
  summarize(count = n())

dtText <- data.frame(
  label = temp$count,
  assessAva = c(0, 0, 0, 0, 1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3),
  treatAva = c(0, 1, 2, 3, 0, 1, 2, 3, 0, 1, 2, 3, 0, 1, 2, 3)
)
 
sgraph3  +
  geom_text(data = dtText, mapping = aes(x = 3, y = 0.75, label = label))
```

## Some Concerning Sample Sizes
We see that several combinations have very low sample sizes, especially at high levels of assessment and low levels of treatment and vice versa, as schools that do not have much assessment resources available tend to not be able to fund or support high treatment. Therefore, we see that outside of our three main diagonals (going from the top left to the bottom right), we have low sample sizes.

Therefore, we should probably refrain from doing an analysis on a (multiplicative) combination of these two variables due to the high chance of the low sample sizes at the extremes skewing our data.

## Breakdown of the Effectiveness of Assessment and Treatment of Mental Health
```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ treatAva + assessAva, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 obs
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```

  From here, we see that a school having mental health services for treatment is highly related to bullying rate, with a p value of about 0.0159, below our standard threshold of 0.05, while assessment is not as important. Our data suggests that the key factor in decreasing bullying rates is having a place where students can go for mental health issues - when students feel their mental health is of a concern, and the school administration works to ensure that students can get the mental health treatment they deserve, bullying rates drop.
  
  We repeat the same process, isolating our variable. 
  
## Breakdown of the Effectiveness of Mental Health Treatment Programs on Bullying Rates
```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ treatAva, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 observations
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```
  
  We get a highly significant p-value, suggesting having available treatment is a key factor in reducing bullying rates.
