---
title: "Mental Health and its Impact on Bullying in Schools"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
---
# Research Question

  While mental health has always been a factor in our education, it has only been recently that it has risen to the forefront of conversation about the impact it has on the education and school environments. In order to capture the impact mental health has on our educational systems today, we look at the mental health of several children in US high school level ages, through bullying rates in schools. We choose to look at bullying rate as it is a common and accurate predictor of general mental health in schools (Schneider, O'Donnell, Stueve, & Coulter), while being easier for an administration to measure as survey respondents may not always be forthcoming.

  We try to depict an accurate picture of the terrible condition of mental health in today's society and answer the following question: What can schools do to effectively decrease bullying rates, and subsequently improve general mental health?

# Initial Setup
```{r}
rm(list = ls())
```

## Libraries
```{r}
library(sas7bdat)
library(tidyverse)
library(ggplot2)
library(data.table)
library(mosaic)
library(ggformula)
library(MASS)
library(dplyr)
library(sjstats)
library(pwr)
library(grid)
library(gtable)
```

## Datasets
```{r}
schooldataraw <- read.sas7bdat("C:\\Users\\julia\\OneDrive\\Desktop\\Mental-Health-In-Schools\\2018-2019 School Survey on Crime and Safety.sas7bdat")
studenthealthraw <- read.csv("C:\\Users\\julia\\OneDrive\\Desktop\\Mental-Health-In-Schools\\Global Student Health Survey.csv")
```

## Functions
```{r}
#functions used with lapply to clean data
convertBoolean <- function(x) {
  return(x==1)
}

convertString <- function(x) {
  riskList <- c("Did not have close friends", "Most of the time or always felt lonely", "Most of the time or always were so worried about something that they could not sleep at night", "Seriously considered attempting suicide", "Made a plan about how they would attempt suicide", "Attempted suicide")
  riskListShort <- c("No close friends", "Always lonely", "Worried to insomnia", "Considered suicide", "Planned suicide", "Attempted suicide")
  
  return(riskListShort[match(x, riskList)])
}
```

## Palettes Used
```{r}
#using a regular color palette in order to help with continuity and not default to blinding neon
threeColorPalette <- c("red3", "gold", "forestgreen")
fiveColorPalette <- c("red3", "orange", "gold", "forestgreen", "royalblue")
sevenColorPalette <- c("red3", "orange", "gold", "forestgreen", "royalblue", "purple2", "pink1")
```

# Data Wrangling and Cleaning

## Cleaning the 2018-2019 School Survey on Crime and Safety
```{r}
#selecting and cleaning up the chosen variables
schooldata <-
  schooldataraw %>%
  dplyr::select(C0174, C0662, C0664, C0666, C0668, C0670, C0672, C0674, C0676, C0678, C0670, C0672, C0674, C0676, C0678, C0680, C0682, C0684, C0686, C0265, C0267, C0273, C0376) %>%
  transmute(id = row_number(), antiBullying = C0174,
         assessAva = 6 - (C0662 + C0664 + C0666), #compiling assessment scores into one score
         treatAva = 6 - (C0668 + C0670 + C0672), #same for treatment
         limitMHProfAccess = C0674, limitFunding = C0676, limitLegal = C0678, limitParental = C0680, 
         limitComm = C0682, limitPaymentPol = C0684, limitStigma = C0686,
         discCyber = C0265, discBullying = C0267, recogBullying = C0273,
         bullyingRate = C0376)
```

```{r}
#taking binary variables and converting them to booleans
booleanRows = c(2, 12, 13, 14)

boolSchoolData <-
  lapply(schooldata[,booleanRows], convertBoolean)

schooldata[ , 2  ] <- boolSchoolData[1]
schooldata[ , 12 ] <- boolSchoolData[2]
schooldata[ , 13 ] <- boolSchoolData[3]
schooldata[ , 14 ] <- boolSchoolData[4]
```

```{r}
#creating a new data table with just limitations
limitations <- data.table(comm = schooldata$limitComm, 
                          funding = schooldata$limitFunding, 
                          legal = schooldata$limitLegal, 
                          MHProfAccess = schooldata$limitMHProfAccess, 
                          parental = schooldata$limitParental, 
                          paymentPol = schooldata$limitPaymentPol, 
                          stigma = schooldata$limitStigma)
```

## Cleaning the Global Student Health Survey
```{r}
studenthealth <-
  studenthealthraw %>%
  transmute(Topic, Description, Question = Greater_Risk_Question, RespPct = Greater_Risk_Data_Value, Sample_Size, Sex, Age) %>%
  filter(Topic == "Mental Health" & Sex != "Total" & 
           Age != "Total" & Age != "13-15" & Age != "13-17") %>%
  na.omit()

studenthealth <-
  studenthealth %>%
  group_by(Age, Sex, Question) %>%
  mutate(SampleSize = sum(Sample_Size)) %>%
  mutate(AffRespPct = sum((Sample_Size)*(RespPct))/SampleSize) %>%
  dplyr::select(Sex, Age, Question, AffRespPct, SampleSize) %>% 
  unique() %>%
  arrange(Sex, Age)

studenthealth$Age = as.numeric(levels(studenthealth$Age))[studenthealth$Age]

studenthealth <- studenthealth
studenthealth$Question = lapply(studenthealth$Question, convertString)
studenthealth <- transform(studenthealth, Question = unlist(Question))
```

# Codebooks

## 2018-2019 School Survey on Crime and Safety
| Variable Name | Description |
| --- | --- |
| ID | The ID number of the School, given for purposes of the study |
| antiBullying | A boolean describing whether a school has anti-bullying programs put into place. |
| assessAva | An aggregate describing whether a school has mental health services for diagnostic assessment available, with a score of 3 being the highest. |
| treatAva | An aggregate describing whether a school has mental health services for treatment available, with a score of 3 being the highest. |
| limitMHProfAccess | A score describing whether inadequate access to mental health professionals was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitFunding | A score describing whether inadequate funding was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitLegal | A score describing whether potential legal issues, such as malpractice, or insufficient supervision were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitParental | A score describing whether lack of parental support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitComm | A score describing whether lack of community support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitPaymentPol | A score describing whether payment polcies were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitStigma | A score describing whether concern for potential stigmatization of the student was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| discCyber | A boolean describing whether school staff were trained in discipline policies and practices related to cyberbullying. |
| discBullying | A boolean describing whether school staff were trained in discipline policies and practices related to bullying. |
| recogBullying | A boolean describing whether school staff were trained in recognizing bullying. |
| bullyingRate | A score describing how often bullying occurs at a school, with 1 being daily and 5 being never. |

## Global Student Health Survey
| Variable Name | Description |
| --- | --- |
| Topic | A variable meant for categorizing, used to filter down to just mental health. |
| Question | A question asked related to mental health, with the relevant time frame being the 12 months before the survey |
| AffRespPct | Percent of people who responded yes to the question. |
| Sample Size | The amount of people who answered this question, with the specified sex and age. |
| Sex | The sex of the recipients. |
| Age | The age of the recipients. |

# The Mental Health Epidemic in Schools

```{r}
studenthealth1 <-
  studenthealth %>%
  mutate(AffResponses = round(AffRespPct * SampleSize / 100)) %>%
  group_by(Question, Sex) %>%
  summarize(AffResponseTotal = sum(AffResponses))
```


```{r}
dtText <- data.frame(
  label = studenthealth1$AffResponseTotal,
  Question = c("No close friends", "No close friends", "Always lonely", "Always lonely", "Worried to insomnia", "Worried to insomnia", "Considered suicide", "Considered suicide", "Planned suicide", "Planned suicide", "Attempted suicide", "Attempted suicide"),
  Sex = rep(c("F", "M"), 6),
  locX = rep(15, 12),
  locY = rep(c(8, 17), 6),
  col = rep(c("red", "blue"), 6)
)

studenthealth$Question = factor(studenthealth$Question, levels = c("No close friends", "Always lonely", "Worried to insomnia", "Considered suicide", "Planned suicide", "Attempted suicide")) #assigning order to facets

p <-
  studenthealth %>%
  ggplot() +
  geom_line(aes(x = Age, y = AffRespPct, color = Sex)) +
  facet_wrap( ~ Question) +
  scale_color_manual(values = c("red2", "red2", "royalblue", "royalblue"), breaks = c("Female", "Male"))

p + 
  geom_text(data = dtText, mapping = aes(x = locX, y = locY, color = col, label = label)) +
  ggtitle("Affirmative Response Count")

```

  Straight percentages are on the y-axis, with sample sizes for the assosciated sex close(ish) to the line. We see the percentage of people who attempt suicide is about 12%, planning is around 14%, and considering is around 16%. More shockingly, these graphics do not include those who have committed suicide, driving these percentages and numbers down, if anything. About one in seven high school aged students have attempted suicide, despite having things we might assume would prevent it, like having close friends.
  
  On a case by case basis, while most people, around 95% consider themselves to have close friends, significant portions of them feel lonely and lost, despite having those close friends. And these preliminary mental health struggles is what turns people to consider suicide, to plan suicide, to attempt suicide, and to commit suicide. Schools must do something to help combat mental health struggles.

## Suicide Rates
```{r}
studenthealth %>%
  filter(Question == "Considered suicide"
         | Question == "Planned suicide"
         | Question == "Attempted suicide") %>%
  ggplot() +
  geom_line(aes(x = Age, y = AffRespPct, linetype = Question, color = Sex)) +
  ggtitle("Suicide Rates, Divided by Sex")
```


# How Effective Are Antibullying Programs
  One of the most commonplace programs put into schools to combat bullying, are the aptly named anti-bullying programs, where teachers are trained to recognize bullying, as well as policies and practices regarding both cyber and physical bullying. We also look at the general response of "anti-bullying" where a school responds affirmatively if there are any anti-bullying programs put into place. 

```{r fig.height=7, fig.width=8}
schoolgraphdata1 <-  
  schooldata %>%
  group_by(discBullying, discCyber) %>%
  summarize(meanRate = mean(bullyingRate))

dummyVerts <- data.frame(discCyber = c(T, F, T, F), discBullying = c(T, T, F, F), 
                         Z =schoolgraphdata1$meanRate) #df with dummy vars at mean rate for each combination

sgraph1 <-
  schooldata %>%
  ggplot(aes(x = bullyingRate)) +
  geom_density(aes(x = bullyingRate, fill = recogBullying, alpha = 0.05)) +
  facet_grid(rows = vars(discBullying), cols = vars(discCyber)) +
  geom_vline(data = dummyVerts, aes(xintercept = Z))

sgraph1 + 
  labs(x = "Trained on Combating Cyberbullying", 
       y = "Trained on Combating Bullying", 
       title = "Bullying Rates With Antibullying Programs Put Into Place") + 
  theme(axis.title.x = element_text(vjust = 205),
        axis.title.y = element_text(angle = 270, vjust = 207))
```

## Bullying Rates for Various Combinations of Anti-Bullying Programs

```{r}
sgraph2 <-
  schooldata %>%
  ggplot(aes(y = bullyingRate, x = recogBullying, color = factor(antiBullying)))

sgraph2 +
  geom_boxplot(lwd = 0.3, color = "black") +
  geom_jitter(alpha = .15) +
  facet_grid(discBullying ~ discCyber) +
  theme(axis.text.x = element_text(vjust = 1)) +
  ggtitle("Bullying Rates for Various Combinations of Anti-Bullying Programs")
```


```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ recogBullying + antiBullying + discCyber + discBullying, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 observations
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```

## Breakdown of the Effectiveness of Anti-Bullying Programs
  After running an ordered logistic regression model, we can see that none of these p values for our predictors are particularly significant, suggesting that training school staff in anti-bullying programs, recognizing bullying, and training them to learn discipline policies for both bullying and cyberbullying are all ultimately ineffective, and these policies don't create a significant impact on bullying rates.

  Therefore, if all of these commonplace school anti-bullying programs are ineffective, what programs can schools put into place to create a significant drop in bullying rates? A good place to start looking is seeing what schools feel limit them from helping students out on the mental health spectrum.

  Sidenote: All of the differences between our levels for bullyingRates are significant, especially for the extremes, 1 and 5, which is reassuring.

## Which limitations are most common to schools?
```{r}
limitationsLong <- pivot_longer(limitations, everything(), names_to = "limType")

limitationsLong %>%
  group_by(limType, value) %>%
  summarize(valCount = n()) %>%
  ggplot(aes(x = limType, y = valCount)) + 
  geom_col(aes(x = limType, y = valCount, fill = as.character(value))) +
  scale_fill_manual(values = threeColorPalette) +
  ggtitle("Common Limitations") +
  guides(fill = guide_legend(title = "Limitation Levels"))
```

```{r}
limBully = data.table(limitations, schooldata$bullyingRate)
limBully <-
  limBully %>%
  rename(bullyingRate = V2)

limitationsLong2 <- pivot_longer(limBully, cols = c("comm", "funding", "legal", "MHProfAccess", "parental", "paymentPol", "stigma"), names_to = "limType") %>%
  transmute(limType = limType, limScore = value, bullyingRate = bullyingRate) 

limitationsLong3 <-
  limitationsLong2 %>%
  group_by(limType, limScore, bullyingRate) %>%
  summarize(bullyingRateCount = n())
```

```{r}
limitationsLong2 %>%  
  ggplot() +
  aes(x = limScore, fill = limType) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = sevenColorPalette) +
  ggtitle("Counts at Various Limitation Types and Scores") +
  guides(fill = guide_legend(title = "Limitation Type"))
```

## Breakdown of What Do Schools Feel They Need and What Do Schools Actually Need

  Although, we might say some combinations like Funding being a very common highly limiting factor jump out to us immediately, we should confirm that it does in fact have a significant impact on bullying rates at schools, as well as any others which might have a noticeable impact. We do this through taking an analysis of variance of the different mean bullying rates for different limitation types in order to see which limitation type has the most impact on bullying rates.
  
  We do this by looking at different combinations of limitation score and limitation type. We plan to do this to answer the following key research question.
 
### How effective are different limitations/resources at combating bullying in schools?
```{r}
limTypes = c("comm", "funding", "legal", "MHProfAccess", "parental", "paymentPol", "stigma")
limPVals = c()
limCohens = c()

for(type in limTypes) {
  limitationsLongTemp <-
    limitationsLong2 %>%
    filter(limType == type)
  
   pVal <- summary(aov(bullyingRate ~ limScore, data = limitationsLongTemp))[[1]][["Pr(>F)"]][[1]]
   limPVals = c(limPVals, pVal)
   
   limCohen <- cohens_f(aov(bullyingRate ~ limScore, data = limitationsLongTemp))[1,2]
   limCohens = c(limCohens, limCohen)
}

data.table(limTypes, limPVals, limCohens)
```

  We see that each limitation type does have an effect on a school's bullying rate, with the level of limitation effecting a school's bullying rate. However, we must look at whether the effects of these limitations are large enough to matter. 
  
  Using Cohen's F statistic, we get Cohen's F values in a range from about 0.077 to 0.127. According to Cohen's suggested Cohen's F Values of 0.1, 0.25, and 0.4 of small, medium, and large effect sizes respectively, we see that all of our effect sizes are in the "small" range, suggesting that these limitations have not much effect on bullying rates, despite their significance.
  
  We can see a graph of how bullying rates differ at different combinations of limitation type and level below.

## Straight Counts of Occurences of Bullying Rates at Various Levels of Specific Limitations
```{r}
limitationsLong3 %>%  
  ggplot() +
  aes(x = limType, fill = as.character(bullyingRate), y = bullyingRateCount) +
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) +
  facet_wrap( ~ limScore) +
  scale_fill_manual(values = fiveColorPalette) +
  guides(fill = guide_legend(title = "Bullying Rate")) +
  ggtitle("Counts of Bullying Rate Occurence at Various Levels")
```

## Proportions of Occurences of Bullying Rates at Various Levels of Specific Limitations
```{r}
temp <-
  limitationsLong3 %>%
  group_by(limType, limScore) %>%
  summarize(totCount = sum(bullyingRateCount))

limitationsLong3 %>%
  left_join(temp, by = c("limType", "limScore")) %>%
  mutate(propRate = bullyingRateCount/totCount) %>%
  ggplot() +
  aes(x = limType, fill = as.character(bullyingRate), y = propRate) +
  geom_bar(position = "stack", stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) +
  facet_wrap( ~ limScore) +
  scale_fill_manual(values = fiveColorPalette) +
  ggtitle("Proportions of Bullying Rates at Various Levels of Limitations")
```

  As we can see, although we see what could be a very slight increase in the proportion of schools with lower bullying rate at higher limitation levels, we see it's not a very large and distinct increase, mirroring what Cohen's F statistic suggests. 

  Clearly, several of the most common anti-bullying programs and styles are all ineffectual, and the common limitations that schools frequently cite turn out not to have too much impact at all. This seems to spell out a grim future for the next generation of students, with bullying remaining a prevalent problem throughout all of our schools, with all of the most common methods against them being ultimately ineffectual and our schools asking for resources with virtually no impact. 
  
  Therefore, where should we go to find a solution to the bullying epidemic in our schools? Following our link between mental health and bullying, we move on to:
  
# Mental Health & Bullying: How Does Assessment and Treatment of Mental Health Impact Bullying
```{r}
sgraph3 <-
  schooldata %>%
  ggplot(aes(x = bullyingRate)) +
  geom_density(aes(x = bullyingRate)) +
  geom_rect(fill = "blue", alpha = 0.003, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf)) +
  facet_grid(rows = vars(assessAva), cols = vars(treatAva), labeller = label_both)

temp <-
  schooldata %>%
  group_by(assessAva, treatAva) %>%
  summarize(count = n())

dtText <- data.frame(
  label = temp$count,
  assessAva = c(0, 0, 0, 0, 1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3),
  treatAva = c(0, 1, 2, 3, 0, 1, 2, 3, 0, 1, 2, 3, 0, 1, 2, 3)
)
 
sgraph3  +
  geom_text(data = dtText, mapping = aes(x = 3, y = 0.75, label = label)) +
  ggtitle("Density Distributions of Bullying Rate at Different Combinations of Treatment and \nAssessment Levels, along with Associated Sample Sizes")
```

## Some Concerning Sample Sizes
We see that several combinations have very low sample sizes, especially at high levels of assessment and low levels of treatment and vice versa, as schools that do not have much assessment resources available tend to not be able to fund or support high treatment. Therefore, we see that outside of our three main diagonals (going from the top left to the bottom right), we have low sample sizes.

Therefore, we should probably refrain from doing an analysis on a (multiplicative) combination of these two variables due to the high chance of the low sample sizes at the extremes skewing our data.

## Breakdown of the Effectiveness of Assessment and Treatment of Mental Health
```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ treatAva + assessAva, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 obs
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```

  From here, we see that a school having mental health services for treatment is highly related to bullying rate, with a p value of about 0.0159, below our standard threshold of 0.05, while assessment is not as important. Our data suggests that the key factor in decreasing bullying rates is having a place where students can go for mental health issues - when students feel their mental health is of a concern, and the school administration works to ensure that students can get the mental health treatment they deserve, bullying rates drop.
  
  We repeat the same process, isolating our variable. 
  
## Breakdown of the Effectiveness of Mental Health Treatment Programs on Bullying Rates
```{r}
orderedLogRegModel <- polr(as.factor(bullyingRate) ~ treatAva, data = schooldata, Hess = T)

coefTable <- coef(summary(orderedLogRegModel))
#technically only true at infinite df, but decent approximation at 2100 observations
pVal <- pnorm(abs(coefTable[, "t value"]), lower.tail = FALSE) * 2


coefTable <- cbind(coefTable, "p value" = pVal)
ci <- confint(orderedLogRegModel)

#printing out confidence intervals and the coefficient table
ci
coefTable
```
  
  We get a highly significant p-value, suggesting having available treatment is a key factor in reducing bullying rates.
 
### Odds Ratio for Levels of Availability for Mental Health Treatment 
```{r}
exp(coefTable[1,1])
```

  The odds ratio is 0.8390, suggesting that for every additional level of treatment availability (which ranges from 0 to 3), the odds of having a higher level of bullying rate decrease by a factor of 0.8390. 
  
## Considerations

  However, while we have seen mental health treatment programs have a highly significant and effective effect on bullying rates, mental health treatment programs are not available to many schools and educational systems across the country, as they require several things - funding, mental health professionals, community and familial support - all things schools seemingly recognize they need.

  But from the apparent differences between the significance and effectiveness of the limitation levels, we can see that even schools with access to these resources, still do not have effective drops in their bullying rates, suggesting these resources are being funnelled into programs with minimal impact. For those educational entities with the necessary resources, we strongly encourage putting in significant investment in concrete mental health treatment programs, rather than wasting resources on programs that have been found to be ineffective, such as mere posters or training teachers. 
  
  However, we recognize that not all schools have access to these resources, and cheaper alternatives must continue to be found. Further work could be done with a cost analysis of each of these mental health programs, along with their subsequent impact, to identify exactly how much each dollar contributes when granted to different programs. By finding the best combination of funding for programs, this would truly address the problem for all schools, not just those with adequate funding. 
  
  While such an analysis would be optimal, the data collection would be much larger by several factors, as it would require schools to disclose budgets and their breakdown, and maintaining accurate breakdowns would be a highly tedious process.
  
  
## Works Cited
  Schneider, Shari Kessel, O'Donnell, Lydia, Stueve, Ann, Coulter, Robert W. S. “Cyberbullying, School 
    Bullying, and Psychological Distress: A Regional Census of High School Students”, American Journal of 
    Public Health 102, no. 1 (January 1, 2012): pp. 171-177.
  