---
title: "Mental Health and its Impact on Bullying in Schools"
output: html_notebook
---


## Initial Setup
```{r}
rm(list = ls())
```

### Libraries
```{r}
library("sas7bdat")
library("tidyverse")
library("ggplot2")
```

### Functions
```{r}
convertBoolean <- function(x) {
  return(x==1)
}
```

### Datasets
```{r}
schooldataraw <- read.sas7bdat("C:\\Users\\julia\\OneDrive\\Desktop\\2018109\\07_FINAL PU DATA_February 5, 2018\\pu_ssocs16.sas7bdat")
```

## Data Wrangling and Cleaning
```{r}
#selecting and cleaning up the chosen variables
schooldata <-
  schooldataraw %>%
  select(C0174, C0662, C0664, C0666, C0668, C0670, C0672, C0674, C0676, C0678, C0670, C0672, C0674, C0676, C0678, C0680, C0682, C0684, C0686, C0265, C0267, C0273, C0376) %>%
  transmute(id = row_number(), antiBullying = C0174,
         assessAva = 6 - (C0662 + C0664 + C0666),
         treatAva = 6 - (C0668 + C0670 + C0672),
         limitMHProfAccess = C0674, limitFunding = C0676, limitLegal = C0678, limitParental = C0680, 
         limitComm = C0682, limitPaymentPol = C0684, limitStigma = C0686,
         discCyber = C0265, discBullying = C0267, recogBullying = C0273,
         bullyingRate = C0376)
```

```{r}
booleanRows = c(1, 11, 12, 13, 14)

boolSchoolData <-
  lapply(schooldata[,booleanRows], convertBoolean)

schooldata[ , 1  ] <- boolSchoolData[1]
schooldata[ , 11 ] <- boolSchoolData[2]
schooldata[ , 12 ] <- boolSchoolData[3]
schooldata[ , 13 ] <- boolSchoolData[4]
schooldata[ , 14 ] <- boolSchoolData[5]
```

## Codebook
| ID | The ID number of the School, given for purposes of the study |
| antiBullying | A boolean describing whether a school has anti-bullying programs put into place. |
| assessAva | An aggregate describing whether a school has mental health services for diagnostic assessment available, with a score of 3 being the highest. |
| treatAva | An aggregate describing whether a school has mental health services for treatment available, with a score of 3 being the highest. |
| limitMHProfAccess | A score describing whether inadequate access to mental health professionals was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitFunding | A score describing whether inadequate funding was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitLegal | A score describing whether potential legal issues, such as malpractice, or insufficient supervision were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitParental | A score describing whether lack of parental support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitComm | A score describing whether lack of community support was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitPaymentPol | A score describing whether payment polcies were a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| limitStigma | A score describing whether concern for potential stigmatization of the student was a limiting factor in the school's efforts to provide mental health services to its students, with 1 being major limitations, and 3 being no limitations. |
| discCyber | A boolean describing whether school staff were trained in discipline policies and practices related to cyberbullying. |
| discBullying | A boolean describing whether school staff were trained in discipline policies and practices related to bullying. |
| recogBullying | A boolean describing whether school staff were trained in recognizing bullying. |
| bullyingRate | A score describing how often bullying occurs at a school, with 1 being daily and 5 being never. |





